Bibat: Batteries-include Bayesian Analysis Template

### 2024-01-30 paper/main.qmd
Update paper

---
title: "Bibat: Batteries-include Bayesian Analysis Template"
abstract: >
  Bayesian statistical workflow offers a powerful way to learn from data,
  but software software projects that implement complex Bayesian workflows
  in practice are unusual, partly due to the difficulty of orchestrating
  Bayesian statistical software. Bibat addresses this challenge by providing a
  full-featured, scalable Bayesian statistical analysis project
  using an interactive template. Bibat is available on the Python Package index,
  documented at <https://bibat.readthedocs.io/> and developed at
  <https://github.com/teddygroves/bibat/>. This paper explains the motivation
  for bibat, briefly describes intended usage, discusses key design choices,
  and reviews several examples of bibat's use in scientific applications.
number-sections: true
author:
  - name: "Teddy Groves"
    orcid: 0000-0002-7109-3270
    email: tedgro@biosustain.dtu.dk
    affiliations:
      - id: 1
        name: Danish Technical University
        department: The Novo Nordisk Foundation Center for Biosustainability
        city: Kongens Lyngby
        country: Denmark
    attributes:
      corresponding: true

bibliography: bibliography.bib

format:
  html:
    embed-resources: true
  acm-pdf:
    keep-tex: true

acm-metadata:
  # comment this out to make submission anonymous
  # anonymous: true

  # comment this out to build a draft version
  final: true

  # comment this out to specify detailed document options
  acmart-options: sigconf, review

  # acm preamble information
  copyright-year: 2024
  acm-year: 2024
  copyright: acmcopyright
  # doi: XXXXXXX.XXXXXXX
  conference-acronym: "KDD"
  conference-name: "KDD"
  conference-date: Aug 25--29, 2024
  conference-location: Barcelona, Esp
  keywords:
    - Python
    - Bayesian workflow
    - Methodology
  teaser:
    image: img/workflow.png
    caption: |
      Schematic representation of a Bayesian workflow implemented using bibat.
      The author inspects their analysis's results, edits files corresponding
      to the boxes on the left, runs the command `make analysis`, then repeats.
      Note that this workflow is modular, accommodates plurality and results in
      a final analysis that can be fully reproduced using a single command.
    description: "DESCRIPTION"

---

@@

Bibat can be installed on the Windows, linux or macos command line on by running
the command `pip install bibat` and then used by running the command `bibat`.
This command triggers an interactive form which prompts the user to select a
range of customisation options. Bibat then creates a new directory containing
code that implements an example analysis, with customisations reflecting the
user's choices. This analysis works immediately, and can be reproduced with
the single command `make analysis` without the need for any further action by
the user: in this sense bibat comes with "batteries included".

@@

Bibat's design was informed by these considerations.

## Accommodate a wide range of statistical analyses

Bibat projects explicitly allow for plurality at the level of input datasets,
prepared datasets, statistical models, fitting modes, computation methods and
analyses. In addition, bibat ensures that there are minimal restrictions on the
kind of components: for example, datasets need not be singular or tabular, and
statistical models need not be representable in formula syntax.

Thanks to these accommodations a project team using bibat should typically not
need to foresee the ultimate requirements of their analysis before starting
the project.

## Encourage reproducibility

Bibat provides a preconfigured makefile with a target `analysis` triggering
creation of an isolated environment, installation of dependencies, data
preparation, statistical computation and analysis of results. In this way a
bibat analysis can be reproduced on most platforms using a single command.

Bibat also provides its Python code in the form of a package configured using
modern conventions for specifying dependencies and configuring tooling, so that
it is easy to maintain reproducibility as the analysis develops.

## Use widely-adopted, open source and actively developed tools

Bibat projects are is written in modern Python and uses pydantic
@pydanticdevelopersPydantic2022 and pandera [@niels_bantilan-proc-scipy-2020]
for data modelling, Stan [@carpenterStanProbabilisticProgramming2017] for
statistical inference, cmdstanpy [@standevelopmentteamCmdStanPy2022] for
python-Stan interface, arviz [@kumarArviZUnifiedLibrary2019] for storing and
analyzing inferences and sphinx [@georgbrandlandthesphinxteamSphinx2022] and
quarto [@Allaire_Quarto_2022] for documentation.

Bibat itself is also written in modern Python and uses the popular
tools cookiecutter [@greenfeldCookiecutter2021], pydantic and click
[@clickdevelopersClickPythonComposable2022].

Bibat's makefile detects the current operating system and attempts to install
cmdstan appropriately if necessary. This functionality addresses a common
issue where researchers find it difficult to install Stan, especially on Windows.

## Implement community standards and best practices for collaborative software development

Bibat projects include a preconfigured test environment, continuous integration,
linting and pre-commit hooks, making them suitable for collaborative software
development. In addition, including documentation as a first class component of
the analysis addresses a common problem in academic statistics projects where
the paper gets out of sync with the code.

Bibat itself is continuously tested to ensure that it works on the operating
systems Linux, macOS and Windows. Bibat's continuous integration runs a test
suite as well as an end-to-end functional test on all supported Python versions.

Bibat is part of the PyOpenSci ecosystem, allowing for community help with
maintenance as well as peer review for code and documentation quality, usability
and accessibility. The PyOpenSci peer review for bibat can be found here: <https://github.com/pyOpenSci/software-submission/issues/83>

# Challenges

This section describes some specific challenges that often affect Bayesian
workflow projects and how bibat addresses them.

## Address complexity using a modular, file based approach

As discussed in @gelmanBayesianWorkflow2020, Bayesian workflows are complicated,
featuring plurality, cyclicity and complexity at many levels. Bibat accommodates
this by separating non-interacting analysis components into modules and by
serialising data to files wherever possible. Prepared datasets, statistical
models, inference configurations, inference results, plots and analyses all
have file representations. Fitting modes, data manipulations and data models are
modularised in code through the use of appropriately structured data classes
and functions.

Thanks to this approach it is possible to perform small sub analyses
individually and to iteratively add components without needing to consider
everything at once.

## Fitting modes

As part of a Bayesian workflow it can be necessary to fit a model and dataset
in different ways. For example, one might perform MCMC sampling of both the
prior and posterior distributions, perform multiple leave-out-one-fold fits for
cross-validation or need to compare MCMC sampling with an optimisation-based
alternative.

Bibat accommodates this need by introducing an abstraction called "fitting
mode". This abstraction allows bibat projects to handle fitting a model
and dataset in different ways appropriately and flexibly. For example, the
provided prior sampling fitting mode creates a Stan input dictionary with the
`likelihood` data variable set to `0`, performs MCMC sampling and writes data
to the `InferenceData` group `prior`. Bibat provides fitting modes corresponding
for prior sampling, posterior sampling and k-fold posterior sampling. Users can
easily add additional fitting modes by copying these examples.

# Post launch use

Bibat has been used for several real statistical analyses.

@grovesteddyDgfreg2023 used bibat to compare a Bayesian and two non-Bayesian
approaches to modelling a biochemical thermodynamics dataset. Bibat facilitated
this analysis even though it was not very large by providing the fitting mode
abstraction, which was very useful for comparing the different methods.

In @grovesBaseball2022, Bibat was used to implement a sports analysis involving
two datasets, two models and four inferences, demonstrating that the generalised
Pareto distribution can be used to describe hitting ability in baseball. This
analysis is now included in bibat as an illustration, along with an accompanying
tutorial.

![](img/baseball.png){width=50%}

In @grovesteddySphincter2024, bibat was used to implement an analysis of
cerebrovascular data from mice, involving two raw datasets, 6 prepared
datasets, 15 models and 15 inferences. In this case bibat's modular design made
it straightforward to iteratively add data transformations and models while
maintaining reproducibility.

Bibat has a growing user community, with 16 GitHub stars at the time of writing.

@@

# Statements and Declarations

The authors have no financial or non-financial interests that are directly or
indirectly related to the work submitted for publication.

# References

>>>>

---
title: "Bibat: Batteries-include Bayesian Analysis Template"
abstract: >
  Bayesian statistical workflow offers a powerful way to learn from data,
  but software software projects that implement complex Bayesian workflows
  in practice are unusual, partly due to the difficulty of orchestrating
  Bayesian statistical software. Bibat addresses this challenge by providing
  a full-featured, scalable Bayesian statistical analysis project using
  an interactive template. Bibat is available on the Python Package index,
  documented at <https://bibat.readthedocs.io/> and developed at <https://github.com/teddygroves/bibat/>. This paper explains the motivation for bibat,
  describes intended usage, discusses bibat's design, compares bibat with similar software, highlights several examples of bibat's use in science and provides links to community resources associated with bibat.
number-sections: true
author:
  - name: "Teddy Groves"
    orcid: 0000-0002-7109-3270
    email: tedgro@biosustain.dtu.dk
    affiliations:
      - id: 1
        name: Danish Technical University
        department: The Novo Nordisk Foundation Center for Biosustainability
        city: Kongens Lyngby
        country: Denmark
    attributes:
      corresponding: true
bibliography: bibliography.bib
keywords: [Bayesian workflow, Methodology, Software]
format:
  typst: default
  html:
    embed-resources: true
  latex: default
---

@@

Bibat can be installed on the Windows, Linux or macOS command line on by running
the command `pip install bibat` and then used by running the command `bibat`.
This command triggers an interactive form which prompts the user to select a
range of customisation options. Bibat then creates a new directory containing
code that implements an example analysis, with customisations reflecting the
user's choices. This analysis works immediately, and can be reproduced with
the single command `make analysis` without the need for any further action by
the user: in this sense bibat comes with batteries included.

@fig-workflow illustrates the components of a bibat-based Bayesian workflow and
shows how it proceeds: the project team edits the code components, then runs
`make analysis` triggering creation of the result components. After inspecting
these they repeat the process, leading to a cycle that ultimately results in a
complete, easily reproducible analysis.

:::{#fig-workflow}

![](img/workflow.png){width=100%}

Schematic representation of a Bayesian workflow implemented using bibat. The
author inspects their analysis's results, edits code corresponding to the
boxes on the left, runs the command `make analysis`, then repeats. The diagram
illustrates several key features of bibat: inference components are modular and
plural, the overall workflow is iterative and cyclical and the whole analysis
can be executed with a single command.

@@

Bibat's design was informed by the aims to accommodate the many sources of
complexity and in a Bayesian workflow project, to ensure easy reproducibility
and to integrate many open-source, widely-adopted and powerful Bayesian workflow
tools and to encourage collaborative development.

As discussed in @gelmanBayesianWorkflow2020, Bayesian workflows are complicated,
featuring plurality, cyclicity and complexity at many levels. As a specialised
Bayesian workflow template, a key goal for bibat was to manage this complexity.
Bibat achieves this aim by separating non-interacting analysis components into
separate, potentially plural modules and by serialising data to files wherever
possible. Prepared datasets, statistical models, inference configurations,
inference results, plots and analyses all have file representations. Fitting
modes, data manipulations and data models are modularised in code through the
use of appropriately structured data classes and functions. Thanks to this
modular approach it is possible to perform small sub analyses individually
and to iteratively expand the analysis by adding components without needing to
consider everything at once. In addition, bibat ensures that there are minimal
restrictions on the components: for example, datasets need not be singular or
tabular and statistical models need not be representable in formula syntax.
Thanks to these accommodations a project team using bibat should typically not
need to foresee the ultimate requirements of their analysis before starting
the project.

Bibat encourages reproducibility mainly by providing a preconfigured makefile
with a target `analysis` that triggers creation of an isolated environment,
installation of dependencies, data preparation, statistical computation and
analysis of results. In this way a bibat analysis can be reproduced on most
platforms using a single command. In particular, this target attempts to install
cmdstan if necessary, using a recipe tailored to the host operating system.
This functionality addresses a common issue where researchers find it difficult
to install Stan, especially on Windows. A second way in which bibat encourages
reproducibility is by providing its Python code in the form of a preconfigured
package. Thanks to the use of modern conventions for specifying dependencies
and configuring tooling, we expect that this will make bibat analyses easier
for other researchers to work with, thereby facilitating replication while also
making initial development more convenient.

Bibat integrates many widely-adopted, open-source and powerful tools to
implement the components of a Bayesian workflow. Bibat projects are is
written in modern Python and uses pydantic @pydanticdevelopersPydantic2022
and pandera [@niels_bantilan-proc-scipy-2020] for data modelling, Stan
[@carpenterStanProbabilisticProgramming2017] for statistical inference,
cmdstanpy [@standevelopmentteamCmdStanPy2022] for python-Stan interface, arviz
[@kumarArviZUnifiedLibrary2019] for storing and analyzing inferences and sphinx
[@georgbrandlandthesphinxteamSphinx2022] and quarto [@Allaire_Quarto_2022]
for documentation.  Bibat itself is also written in modern Python and uses the
popular tools cookiecutter [@greenfeldCookiecutter2021], pydantic and click
[@clickdevelopersClickPythonComposable2022].

To encourage collaborative development of Bayesian workflow projects, Bibat
projects include a preconfigured test environment, continuous integration,
linting and pre-commit hooks. In addition, including documentation as a
first class component of the analysis addresses a common problem in academic
statistics projects where the paper gets out of sync with the code. Bibat itself
is continuously tested to ensure that it works on the operating systems Linux,
macOS and Windows. Bibat's continuous integration runs a test suite as well as
an end-to-end functional test on all supported Python versions.

# Fitting modes

The most novel part of bibat's design is the introduction of an abstraction
called "fitting mode", which allows bibat project to handle different ways
of fitting a model to a dataset appropriately and flexibly.  This is often
necessary as part of a Bayesian workflow: for example, one might perform
MCMC sampling of both the prior and posterior distributions, perform multiple
leave-out-one-fold fits for cross-validation or need to compare MCMC sampling
with an optimisation-based alternative.

Fitting modes in bibat projects take the form of instances of the class
`FittingMode`. Each fitting mode contains a name, a function that fits a
prepared dataset and instructions for how and where to save the results.
For example, the provided prior sampling fitting mode creates a Stan input
dictionary with the `likelihood` data variable set to `0`, performs MCMC
sampling and writes data to the `InferenceData` group `prior`. This design
allows multiple fits to be stored alongside each other naturally, facilitating
many common Bayesian workflow analyses, and can easily be configured to prevent
unnecessary computations.

Bibat provides fitting modes corresponding for prior sampling, posterior
sampling and k-fold posterior sampling. Users can easily add additional fitting
modes by copying these examples, or modify the `FittingMode` class to achieve
even richer functionality.

@@

# Case studies

The following cases illustrate how bibat has been used in practice to facilitate
Bayesian workflow projects.

@grovesteddyDgfreg2023 used bibat to compare a Bayesian and two non-Bayesian
approaches to modelling a biochemical thermodynamics dataset. Bibat facilitated
this analysis even though it was not very large---the final analysis contained
one dataset, three models and three inferences---because the fitting mode
abstraction allowed for straightforward comparison of the different methods.
Additionally, bibat made it easier to iteratively investigate and discard models
that did not form part of the final analysis.

In @grovesBaseball2022, Bibat was used to implement a sports analysis involving
two datasets, two models and four inferences, demonstrating that the generalised
Pareto distribution can be used to describe hitting ability in baseball. This
analysis is now included in bibat as an illustration, along with an accompanying
tutorial. An illustrative graphic from this analysis is shown in @fig-baseball.

In this case bibat was useful because of its ability to implement arbitrary
statistical models, as latent generalised Pareto distributions are not supported
by generic regression packages like brms and bambi. Further, bibat's modular
design made it easier to implement this medium-sized analysis with two datasets,
two models and six inferences.

![](img/baseball.png){width=100%}

In @grovesteddySphincter2024, bibat was used to implement a large analysis
of cerebrovascular data from mice, involving two raw datasets, 6 prepared
datasets, 15 models and 15 inferences. Again it was not possible to implement
the analysis using formula-based packages due to the need for custom modelling
approaches to take into account expert knowledge about both the target system
and the measurement apparatus. Due to the number of datasets and models that
needed to be considered, implementing this analysis without using bibat or a
similar structured approach would have required a combination of more time and/
or compromises in software quality.

# Community

Bibat is developed in public and encourages community contribution. Please see the contributing page <https://github.com/teddygroves/bibat/blob/main/CONTRIBUTING.rst> and code of conduct <https://github.com/teddygroves/bibat/blob/main/CODE_OF_CONDUCT.md> if you would like to help develop bibat.

Bibat has a growing user community, with 16 GitHub stars at the time of
writing, and is affiliated with cmdstanpy through a link on its documentation
website. Bibat is also affiliated with the Python scientific software community
PyOpenSci, allowing for help with maintenance as well as peer review for
code and documentation quality, usability and accessibility. The PyOpenSci
peer review for bibat can be found here: <https://github.com/pyOpenSci/software-submission/issues/83>.

# Author contributions

# Acknowledgements
